---
title: "Lab 2: Distances and Projections"
subtitle: 'ESS 523c: Environmental Data Science Applications'
author: 
  - name: Megan Hoover
  - email: megan.hoover21@gmail.com
format: html
---

### Lab Intro

Author: Megan Hoover

This lab connects directly to recent policy discussions about border enforcement and the so-called “100-mile” border zone — for example, shifts in migration policy after the end of the pandemic-era Title 42 expulsions in 2023 have renewed attention on how proximity to the border affects enforcement, resource allocation, and civil liberties (see recent coverage: https://www.reuters.com/world/us/us-ceases-using-title-42-pandemic-order-control-migration-2023-05-11/).

More recent operational discussions involving ICE and border enforcement through 2025–2026 have emphasized how interior and border-zone enforcement strategies depend on distance-to-border metrics; readers can consult summaries on civil-liberties implications (ACLU) or policy analyses that discuss the “100-mile” enforcement footprint for background: https://www.aclu.org/know-your-rights/border-zone, https://www.americanimmigrationcouncil.org/blog/border-patrol-charlotte-atlanta-100-mile-zone/.

### Install Libraries

```{r}
pacman::p_load(
  tidyverse, sf, units, tigris, rnaturalearth, gghighlight, ggrepel, flextable, leaflet)

options(tigris_use_cache = TRUE)
```

### **Question 1: Build Your Spatial Dataset**

For this lab, we'll need three datasets:

1.  CONUS state boundaries (1.1)

2.  North American country boundaries — US, Mexico, Canada (1.2)

3.  All US cities (1.3)

#### **1.1 Define a Projection**

Distance calculations require a projected coordinate system that minimizes distance distortion at the scale of CONUS. We will use the **North America Equidistant Conic**:

```{r}
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```

This PROJ string defines:

| Parameter                | Value      | Meaning                       |
|--------------------------|------------|-------------------------------|
| `+proj=eqdc`             | —          | Equidistant Conic projection  |
| `+lat_0=40`              | 40°N       | Latitude of projection center |
| `+lon_0=-96`             | 96°W       | Central meridian              |
| `+lat_1=20`, `+lat_2=60` | 20°N, 60°N | Standard parallels            |
| `+datum=NAD83`           | —          | North American Datum 1983     |
| `+units=m`               | —          | Output in meters              |

Distortion is minimized between the two standard parallels — 20°N and 60°N encompasses all of CONUS, making this projection well-suited for national-scale distance analysis.

#### **1.2 CONUS State Boundaries**

Use `tigris::states()` to get US state boundaries, filter to CONUS, and transform to the equidistant projection.

**\
Tip**

`tigris` is the modern, CRAN-stable replacement for `USAboundaries`. It pulls boundaries directly from the US Census Bureau TIGER/Line files. Load the data with `states(cb = TRUE, progress_bar = FALSE)`. Filter the state codes (`STUSPS` column) to exclude Alaska, Hawaii, and US territories, then apply `st_transform(eqdc)` to reproject.

```{r}
?states
states<-states(cb = TRUE, progress_bar = FALSE) |>
  filter(!STUSPS %in% c("AK", "HI", "AS", "GU", "MP", "PR", "VI","DC")) |>
  st_transform(eqdc)
```

#### **1.3 North American Country Boundaries**

Use `rnaturalearth::ne_countries()` to get country boundaries for the United States, Canada, and Mexico.

**Tip**

`ne_countries()` returns an `sf` object directly when you specify `returnclass = "sf"`. Filter on the `admin` column to keep only the three countries you need, then transform to `eqdc`. See `?rnaturalearth::ne_countries` for available scales and parameters.

```{r}
?rnaturalearth::ne_countries
countries<- ne_countries(returnclass = "sf") |>
filter(admin %in% c("United States of America", "Canada","Mexico"))|>
 st_transform(eqdc) 
```

#### **1.4 US City Locations**

Go to [simplemaps.com/data/us-cities](https://simplemaps.com/data/us-cities) and download the free basic dataset into your `data/` directory.

Read it in with `readr::read_csv()` and explore it — understand what columns are available before proceeding.

The raw data is tabular — coordinates are stored as numeric columns, not geometry. Convert it to a spatial object using `st_as_sf()`, specifying which columns hold X and Y coordinates and what CRS they use. Raw lat/lon from a CSV is almost always WGS84 (EPSG:4326).

**Tip**

After creating the `sf` object, you’ll need to remove rows that fall outside CONUS. The `st_union(states)` geometry provides a boundary to filter against. See `?sf::st_filter` for how to select geometries that intersect or fall within a bounding polygon.

```{r}
?sf::st_filter
cities<- read_csv("data/uscities.csv") |>
  st_as_sf(coords = c("lng", "lat"), crs = 4326) |>
  st_transform(st_crs(states)) |> #Reprojects cities to same CRS as states object
   st_filter(st_union(states)) #Merges all state polygons into one big one & keeps only points intersecting boundary line.
  
```

**Verify all three datasets before proceeding**

Run these checks before moving to Q2 — a CRS mismatch here will silently corrupt every distance calculation downstream.

```{r}
# All three must return the same CRS
st_crs(states) == st_crs(countries)
st_crs(states) == st_crs(cities)


# Row counts should be reasonable
nrow(states)    # should be 48
nrow(cities)    # should be ~28,000

```

### **Question 2: Distance Calculations**

Here we calculate the distance from each US city to four boundaries:

1.  The US national border (coastline + land borders, resolved)

2.  The nearest state border (internal boundaries preserved)

3.  The Mexican border

4.  The Canadian border

In all cases, since we want distance to a **border** (a line), we must cast our polygon geometries to `MULTILINESTRING` before computing distances. The choice of `st_union` vs. `st_combine` determines whether interior boundaries are dissolved or preserved — a distinction that matters for questions 2.1 and 2.2.

**The geometry type determines the answer**

- `POLYGON` / `MULTIPOLYGON` → distance to the interior is **0** (you’re inside the area)

- `MULTILINESTRING` → distance to the **boundary line** — what we actually want

```         
# Wrong: distance to a polygon st_distance(city_point, state_polygon)   # returns 0 if city is inside  # Right: distance to the border st_distance(city_point, state_polygon |> st_cast("MULTILINESTRING"))
```

#### **2.1 Distance to US National Border (km)**

The national border is the **resolved** outer boundary of all states — internal state lines dissolved, only the coastline and land borders remain.

**Tip**

Use `st_union()` to merge all state polygons into a single boundary (dissolving internal borders), then cast to `MULTILINESTRING` so distances are measured to the **border line** rather than interior. Compute distance using `st_distance()` and convert the result to kilometers using `set_units()` from the `units` package — this ensures clean numeric output and correct unit handling downstream.

Store the result as a new column `dist_to_border` in your cities data frame. Produce a **flextable** of the 5 cities farthest from the national border — include city name, state, and distance.

```{r}
?st_distance
?set_units
us_border <- states |>
  st_union() |> # dissolve internal state boundaries
  st_cast("MULTILINESTRING")# convert polygon to boundary lines

cities <- cities |>
  mutate(dist_to_border = st_distance(geometry, us_border),
    dist_to_border = set_units(dist_to_border, "km"))


cities |>
  mutate(dist_to_border = drop_units(dist_to_border)) |> #removes the units, leaving a plain numeric column.
  arrange(desc(dist_to_border)) |>
  slice_head(n = 5) |>
  select(city, state_name, dist_to_border) |>
  flextable() |>
  set_header_labels(
    city = "City",
    state_name = "State",
    dist_to_border= "Distance to Border (km)") |>
  set_caption("Top 5 Cities Farthest From US Borders by")|>
  autofit()
```

#### **2.2 Distance to Nearest State Border (km)**

The nearest state border uses **preserved** internal boundaries — we want every state line, not just the outer edge. Use `st_combine()` which keeps all boundaries, then cast to `MULTILINESTRING`.

Store as `dist_to_state`. Produce a flextable of the 5 cities farthest from any state border.

```{r}
us_border <- states |>
  st_combine() |> # keep internal state boundaries
  st_cast("MULTILINESTRING")# convert polygon to boundary lines

cities <- cities |> #us_border is state border in this
  mutate(dist_to_state = st_distance(geometry, us_border),
    dist_to_state = set_units(dist_to_state, "km"))


cities |>
  mutate(dist_to_state = drop_units(dist_to_state)) |> #removes the units, leaving a plain numeric column.
  arrange(desc(dist_to_state)) |>
  slice_head(n = 5) |>
  select(city, state_name, dist_to_state) |>
  flextable() |>
  set_header_labels(
    city = "City",
    state_name = "State",
    dist_to_state= "Distance to State Border (km)") |>
  set_caption("Top 5 Cities Farthest From US State Borders")|>
  autofit()

```

**Tip**

Think carefully about why the answers for 2.1 and 2.2 differ. The city farthest from the national border might be close to a state line. The city farthest from any state line must be far from all internal AND external borders.

- **2.1 (National border)**: distance only to the outer boundary (coastline + international borders).

  - A city deep inside a state can be far from the national border but close to an internal state line.

<!-- -->

- **2.2 (State borders)** → distance to any state line, including internal boundaries.

```         
-    The farthest city from all state lines must be roughly in the center of a large state.
```

#### **2.3 Distance to Mexico (km)**

Isolate Mexico from your countries object. Cast to `MULTILINESTRING`. Compute distance from each city to the Mexican border. Store as `dist_to_mexico`. Produce a flextable of the 5 cities farthest from Mexico.

```{r}
 notus<- countries |>
      filter(admin == "Mexico") |>
      st_cast("MULTILINESTRING") #convert polygon to boundary lines
  

cities <- cities |> #us_border is state border in this
  mutate(dist_to_mexico = st_distance(geometry, notus),
    dist_to_mexico = set_units(dist_to_mexico, "km"))


cities |>
  mutate(dist_to_mexico = drop_units(dist_to_mexico)) |> #removes the units, leaving a plain numeric column.
  arrange(desc(dist_to_mexico)) |>
  slice_head(n = 5) |>
  select(city, state_name, dist_to_mexico) |>
  flextable() |>
  set_header_labels(
    city = "City",
    state_name = "State",
    dist_to_mexico = "Distance to Mexico Border (km)") |>
  set_caption("Top 5 Cities Farthest From Mexico Border")|>
  autofit()

```

#### **2.4 Distance to Canada (km)**

Same approach as 2.3, for Canada. Store as `dist_to_canada`. Produce a flextable of the 5 cities farthest from Canada.

```{r}
notus<- countries |>
      filter(admin == "Canada") |>
      st_cast("MULTILINESTRING") #convert polygon to boundary lines
  

cities <- cities |> #us_border is state border in this
  mutate(dist_to_canada = st_distance(geometry, notus),
    dist_to_canada = set_units(dist_to_canada, "km"))


cities |>
  mutate(dist_to_canada = drop_units(dist_to_canada)) |> #removes the units, leaving a plain numeric column.
  arrange(desc(dist_to_canada)) |>
  slice_head(n = 5) |>
  select(city, state_name, dist_to_canada) |>
  flextable() |>
  set_header_labels(
    city = "City",
    state_name = "State",
    dist_to_canada = "Distance to Canadian Border (km)") |>
  set_caption("Top 5 Cities Farthest From Canadian Border")|>
  autofit()
```

### **Question 3: Mapping Distance Patterns**

Now visualize the distance data you computed. You will use `ggplot2` for all maps, `ggrepel::geom_label_repel()` for non-overlapping city labels, and `gghighlight` to emphasize subsets.

#### **3.1 Reference Map**

Create a base reference map showing:

- North America (US, Canada, Mexico) as filled grey polygons

- CONUS state boundaries as dashed lines

- The 10 largest US cities by population as labeled points

**Tip**

Use `geom_sf()` for spatial layers. For labels that don’t overlap:

```         
ggrepel::geom_label_repel(   data = big_cities,   aes(geometry = geometry, label = city),   stat = "sf_coordinates",   size = 3 )
```

`stat = "sf_coordinates"` extracts point coordinates from the geometry for label placement — required when using `ggrepel` with `sf` objects.

```{r}
ggplot() +
  # North American countries
  geom_sf(data = countries, fill = "grey80", color = "black") +
   # CONUS state boundaries as dashed lines
  geom_sf(data = states, fill = NA, color = "black", linetype = "dashed") +
   # Plot and label the 10 largest cities inline
  geom_sf( data = cities %>% arrange(desc(population)) %>% slice_head(n = 10),
    color = "red", size = 2 ) +
  ggrepel::geom_label_repel(
    data = cities %>% arrange(desc(population)) %>% slice_head(n = 10),
    aes(geometry = geometry, label = city),
    stat = "sf_coordinates",
    size = 3,
    min.segment.length = 0 ) +
  # Theme and labels
  theme_minimal() +
  labs(
    title = "Reference Map: CONUS and Major US Cities",
    subtitle = "10 Largest US Cities by Population",
    caption = "Data: tigris, rnaturalearth, simplemaps")
```

#### **3.2 Distance to National Border**

Color all US cities by their distance to the national border using a continuous color scale. Highlight and label the 5 cities farthest from the border.

Your map should: - Use `scale_color_viridis_c()` for the distance gradient - Show the CONUS outline - Label the 5 farthest cities with `ggrepel`

```{r}
ggplot() +
  # CONUS outline
  geom_sf(data = states, fill = NA, color = "black") +
  # All US cities colored by distance to national border
  geom_sf(data = cities,
    aes(color = drop_units(dist_to_border)),
    size = 1) +
  scale_color_viridis_c(option = "plasma", name = "Distance to Border (km)") +
   # Label 5 farthest cities inline
  ggrepel::geom_label_repel(data = cities %>%
      mutate(dist_km = drop_units(dist_to_border)) %>%
      arrange(desc(dist_km)) %>%
      slice_head(n = 5),
    aes(geometry = geometry, label = city),
    stat = "sf_coordinates",
    size = 3,
    min.segment.length = 0) +
  # Theme and labels
  theme_minimal() +
  labs(
    title = "US Cities by Distance to National Border",
    subtitle = " Top 5 Farthest Cities Labeled",
    caption = "Data: tigris, rnaturalearth, simplemaps" )
```

In 2–3 sentences:

**(1) where are the cities farthest from the national border?** The cities farthest from the national border are located in Kansas

**(2) Does the geographic pattern make sense given the shape of CONUS?** It makes sense because the projection is equidistant conic, which preserves distance.

**(3) Name one water resources application where distance to a national border would be a meaningful variable**. Distance of aquifers and watershed management to national borders matters in terms of how water is managed.

#### **3.3 Distance to Nearest State Border**

Color all cities by distance to the nearest state border. Highlight and label the 5 cities farthest from any state line.

```{r}
ggplot() +
  # CONUS outline
  geom_sf(data = states, fill = NA, color = "black") +
  
  # Cities colored by distance to nearest state border
  geom_sf(
    data = cities,
    aes(color = drop_units(dist_to_state)),
    size = 1) +
  scale_color_viridis_c(
    option = "plasma",
    name = "Distance to State Border (km)" ) +
  
  # Label the 5 farthest cities inline
  ggrepel::geom_label_repel(
    data = cities %>%
      mutate(dist_km = drop_units(dist_to_state)) %>%
      arrange(desc(dist_km)) %>%
      slice_head(n = 5),
    aes(geometry = geometry, label = city),
    stat = "sf_coordinates",
    size = 3,
    min.segment.length = 0) +
   theme_minimal() +
  labs(
    title = "US Cities by Distance to Nearest State Border", subtitle = "Cities Farther from Any State Line Are Highlighted",
    caption = "Data: tigris, rnaturalearth, simplemaps" )

```

In 2–3 sentences:

**(1) how does the pattern differ from 3.2?** The pattern now shows the five cities closest to the state borders. It's no longer int he middle of the United States, but in the Southern part; Texas.

**(2) Which states appear to have the largest interior areas — do any water resources features (large river basins, aquifer extents) correlate with the states where cities can be far from borders?** States like Texas, Colorado, New Mexico, Utah, and Arizona have the largest interior areas. There are large river systems that go through these states, near the cities that are farthest from the borders. Some of those rivers are the Colorado River, the Arkansas River, The Rio Grande River, The Sacramento River, The Brazos River, and the Gila River.

**(3) What would a hydrologist use this distance for?** A hydrologist could use this distance to determine watershed management oversight, account for evaporation calculations to meet quotas of different compacts, and determine areas where most of the water is being used.

#### **3.4 Equidistance Zone: Mexico and Canada**

Identify cities that are **approximately equidistant** from the Mexican and Canadian borders — within ±100 km of equal distance from both.

**Tip**

Create a variable for the absolute difference between the two distances, then use `gghighlight()` to emphasize cities where that difference is less than 100 km. Label the 5 most populous cities in this zone with `ggrepel`.

```{r}
ggplot() +
  # CONUS outline
  geom_sf(data = states, fill = NA, color = "black") +
  
  # All cities colored by difference in distance
  geom_sf(data = cities %>%
      mutate(diff_km = abs(drop_units(dist_to_mexico - dist_to_canada))),
    aes(color = diff_km),
    size = 1) +
  
  scale_color_viridis_c(
    option = "plasma",
    name = "Distance (km)") +
  
  # Highlight cities within ±100 km difference
  gghighlight::gghighlight(diff_km <= 100) +
  
  # Label 5 most populous cities in the equidistant zone
  ggrepel::geom_label_repel(
    data = cities %>%
      mutate(
        diff_km = abs(drop_units(dist_to_mexico - dist_to_canada)) ) %>%
      filter(diff_km <= 100) %>%
      arrange(desc(population)) %>%
      slice_head(n = 5),
    aes(geometry = geometry, label = city),
    stat = "sf_coordinates",
    size = 3,
    min.segment.length = 0) +
  
  theme_minimal() +
  labs(
    title = "Cities Equidistant from Mexico and Canada (±100 km)",
    caption = "Data: tigris, rnaturalearth, simplemaps" )

```

In 2–3 sentences:

**(1) describe where the equidistant zone runs geographically**

The equidistant zone runs geographically left to right, from the center of California and diagonally down to the upper East Florida coast.

**(2) does it surprise you that this band cuts through the interior of the country where it does?** No it doesn't surprise me after looking where some of the major river systems are. This belt runs roughly between the Missouri and the Arkansas Rivers.

**(3) What does this zone imply about the relative influence of Canadian vs. Mexican climate systems on cities in this corridor?**

This zone may face more climate variability from the the colder moist air from the North and the warmer moist air from the South. This may also influence water availability, which would help explain why those cities are aligned between rivers.

### **Question 4: Real-World Application — The 100-Mile Border Zone**

#### **Background**

Federal regulations give U.S. Customs and Border Protection (CBP) authority to operate within **100 miles of any US “external boundary”** — including both land borders and coastlines. Within this zone, CBP agents can conduct stops and searches without a warrant or reasonable suspicion. Federal courts have repeatedly upheld this authority. Nearly 2 out of 3 people live within the 100 mile border zone.

The American Civil Liberties Union has documented and challenged this policy. Their analysis is [here](https://www.aclu.org/other/constitution-100-mile-border-zone).

100 miles ≈ **160 kilometers**. Your distance calculations are already in kilometers.

#### **4.1 Quantify the Zone**

Using your `dist_to_border` column, calculate:

- How many cities fall within the 100-mile zone?

- What is the total population living in cities within the zone?

- What percentage of the total US city population does this represent?

- Does your result match the ACLU’s published estimate?

Present your findings as a formatted **flextable**.

**Tip**

Filter to cities within 160 km, drop geometry if needed, then summarize: count rows (`n()`), sum population, and compute the percentage relative to total city population. You will need to reference total city population across the *entire dataset*, not just filtered rows — plan your calculation carefully.

```{r}
cities |>
 mutate(dist_to_border= drop_units(dist_to_border)) |>
  summarise(num_cities = sum(dist_to_border<= 160, na.rm = TRUE),
    total_pop = sum(population[dist_to_border <= 160], na.rm = TRUE),
    percent_of_total = (sum(population[dist_to_border<= 160], na.rm = TRUE) / sum(population, na.rm = TRUE)) * 100
  ) |>
  flextable() |>
  set_header_labels(
    num_cities = "Number of Cities",
    total_pop = "Total Population",
    percent_of_total = "Percent of US City Population") |>
  set_caption("Cities and Population Within 100-Mile Border Zone") |>
autofit()
```

#### **4.2 Map the Zone**

Create a map highlighting all cities within the 100-mile zone using `gghighlight`.

- Color cities by distance to the border using a gradient from `"orange"` to `"darkred"`

- Label the **10 most populous** cities in the zone

```{r}
ggplot() +
  # CONUS outline
  geom_sf(data = states, fill = NA, color = "black") +
  
  # All cities colored by distance to border
  geom_sf(data = cities |>
      mutate(dist_to_border = drop_units(dist_to_border)),
    aes(color = dist_to_border),
    size = 1) +
  
  # Orange to dark red gradient
  scale_color_gradient(
    low = "orange",
    high = "darkred",
    name = "Distance to Border (km)") +
  
  # Highlight cities within 100 miles (~160 km)
  gghighlight(dist_to_border <= 160) +
  
  # Label 10 most populous cities in the zone
  ggrepel::geom_label_repel(
    data = cities |>
      mutate( dist_to_border= drop_units(dist_to_border))|> filter(dist_to_border<= 160) |>
      arrange(desc(population)) |>
      slice_head(n = 10),
    aes(geometry = geometry, label = city),
    stat = "sf_coordinates",
    size = 3,
    min.segment.length = 0 ) +
  theme_minimal() +
  labs(
    title = "10 Most populated US Cities Within the 100-Mile Border Zone",
    subtitle = "Highlighted Cities Are Within ~160 km of the National Border",
    caption = "Data: tigris, rnaturalearth, simplemaps" )

```

#### **4.3 Most Populous City per State in the Zone**

Repeat the map from 4.2, but instead of the 10 most populous overall, label the **most populous city in each state** that falls within the zone.

**Tip**

filter(cities, as.numeric(dist_to_border) \< 160) \|\> group_by(state_name) \|\> slice_max(population, n = 1)

```{r}
ggplot() +
  # CONUS outline
  geom_sf(data = states, fill = NA, color = "black") +
  
  # All cities colored by distance to border
  geom_sf(data = cities |>
      mutate(dist_to_border = drop_units(dist_to_border)),
    aes(color = dist_to_border),
    size = 1) +
  
  # Orange dark red gradient
  scale_color_gradient(
    low = "orange",
    high = "darkred",
    name = "Distance to Border (km)") +
  
  # Highlight cities within 100 miles (~160 km)
  gghighlight(dist_to_border <= 160) +
  
  # Label most populous city per state within the zone
  ggrepel::geom_label_repel(data = cities |>
      mutate(dist_to_border = drop_units(dist_to_border))|>
      filter(dist_to_border <= 160) |>       # keep only cities in the zone
      group_by(state_name) |>         # group by state
      slice_max(order_by = population, n = 1, with_ties = FALSE) |> # top city per state
      ungroup(),
    aes(geometry = geometry, label = city),
    stat = "sf_coordinates",
    size = 3,
    min.segment.length = 0) +
  
  theme_minimal() +
  labs(
    title = "Most Populous Border-Zone City per State",
    subtitle = "Cities Within ~160 km of the National Border",
    caption = "Data: tigris, rnaturalearth, simplemaps")

```

In 2–3 sentences:

**(1) how many states have at least one city in the zone?** There are 35 states that have at least one city in the zone.

```{r}
cities |>
  mutate(dist_to_border = drop_units(dist_to_border)) |> # convert units to numeric
  filter(dist_to_border <= 160) |>                       # keep only cities in the 100-mile zone
  distinct(state_name) |>   # get unique states
  nrow()     
```

**(2) Are there states where the most populous city is NOT in the zone — what does that tell you about that state’s geography?** Yes, there are many states without the most populous city not in the zone. Some of those cities are Denver, Salt Lake City, Albuquerque, Wichita, Cheyenne, etc.

```{r}
cities |>
  mutate(dist_to_border = drop_units(dist_to_border)) |>
  group_by(state_name) |>
  slice_max(order_by = population, n = 1, with_ties = FALSE) |>  # top city per state
  ungroup() %>%
  filter(dist_to_border > 160) %>%  # most populous city not in 100-mile zone
  select(state_name, city, dist_to_border) %>% 
  arrange(desc(dist_to_border))

```

**(3) How might the jurisdictional boundary created by this rule create challenges for environmental monitoring programs that cross state or national boundaries?** Crossborder coordination becomes harder because regulatory zones may not match natural water system boundaries, leading to gaps in pollution monitoring, water resource management, or disaster response.

### **Question 5: Interactive Leaflet Exercise — Build an interactive map**

This hands-on exercise ties the distance calculations and mapping skills from Q1–Q4 into a single interactive web map. The goal is a compact, well-documented Leaflet map with layer controls, informative popups, and two toggleable thematic layers: (A) the ±100 km equidistance band (Mexico vs Canada) and (B) the 100-mile (≈160 km) national-border zone.

Learning objectives:

- Practice turning `sf` results into interactive HTML widgets with `leaflet`.

- Build informative popups that show computed distances and metadata.

- Implement layer groups and a legend so users can toggle overlays.

- Consider performance and projection issues when publishing web maps.

#### **Deliverables**

Create an interactive `leaflet` map that includes:

- **Base layers**: CONUS state boundaries and at least one provider tile basemap (e.g., CartoDB, OpenStreetMap)

- **Data layers** (toggleable via `addLayersControl`):

  - All cities colored by distance to the national border (use a color palette like `colorNumeric("viridis")`; see `?leaflet::colorNumeric`)

  - Cities in the equidistance zone (±100 km Mexico/Canada difference) highlighted separately

  - Cities in the 100-mile border zone highlighted separately

- **Popups**: When you click a city marker, display relevant information: city name, state, population, and distances to borders

- **Legend**: Show what the color scale represents

**Tip**

Key `leaflet` functions: - `leaflet()` to initialize - `addProviderTiles()` for basemaps - `addCircleMarkers()` or `addMarkers()` for point data - `addPolygons()` for state boundaries - `addLegend()` for a color scale legend - `addLayersControl()` to make layers toggleable

See `?leaflet` and the [Leaflet for R documentation](https://rstudio.github.io/leaflet/) for syntax and options. Remember that `leaflet` works with WGS84 coordinates (EPSG:4326), so you may need to transform your data before plotting.

```{r}
# create a simple color palette
pal <- colorNumeric(
  "viridis",
  domain = cities |> mutate(dist_to_border = drop_units(dist_to_border)) |> pull(dist_to_border))

leaflet() |>
  # basemap
  addProviderTiles(providers$OpenStreetMap) |>
  # state boundaries
  addPolygons(
    data = st_transform(states, 4326),
    color = "black",
    weight = 1,
    fill = FALSE) |>
  
  # all cities (colored by distance)
  addCircleMarkers(
    data = cities |>
      st_transform(4326) |>
      mutate(
        dist_to_border = drop_units(dist_to_border),
        dist_to_mexico = drop_units(dist_to_mexico),
        dist_to_canada = drop_units(dist_to_canada)),
    
    color = ~pal(dist_to_border),
    radius = 2,
    fillOpacity = 0.3,
    popup = ~paste0(
      city, ", ", state_name, "<br>",
      "Pop: ", population, "<br>",
      "Border: ", round(dist_to_border, 1), " km<br>",
      "Mexico: ", round(dist_to_mexico, 1), " km<br>",
      "Canada: ", round(dist_to_canada, 1), " km"),
    
    group = "All Cities") |>
  
  # 100-mile zone
  addCircleMarkers(
    data = cities |>
      st_transform(4326) |>
      mutate(dist_to_border = drop_units(dist_to_border))|>
      filter(dist_to_border <= 160),
    color = "red",
    radius = 2,
     fillOpacity = 0.3,
    popup = ~city,
    group = "100-mile zone") |>
  
  # equidistant zone
  addCircleMarkers(
    data = cities |>
      st_transform(4326) |>
      mutate(diff = abs(drop_units(dist_to_mexico - dist_to_canada))) |>
      filter(diff <= 100),
    color = "blue",
    radius = 2,
     fillOpacity = 0.3,
    popup = ~city,
    group = "Equidistant zone") |>
  
  # legend
  addLegend(
    "bottomright",
    pal = pal,
    values = cities |> mutate(dist_to_border = drop_units(dist_to_border)) |> pull(dist_to_border),
    title = "Distance to border (km)") |>
  
  # layer control
  addLayersControl(
    overlayGroups = c("All Cities", "100-mile zone", "Equidistant zone"),
    options = layersControlOptions(collapsed = FALSE) )
```

**Methods note**: **In 2–3 sentences, describe the CRS transformation needed to convert from `eqdc` to `leaflet`-compatible coordinates, and any design choices you made to keep the map responsive (e.g., filtering, clustering, or simplifying geometries).**

The spatial data were originally projected in an Equidistant Conic projection (EPSG defined by `eqdc`) to ensure accurate distance calculations across CONUS. Because Leaflet requires geographic coordinates in WGS84, all spatial layers were transformed using `st_transform(4326)` before mapping. Each type of city marker was assigned to a different group to allow turning on and off layers. To maintain performance, the map uses small circle markers, filtered subsets for highlighted layers, and avoids rendering unnecessary attributes or overly large geometries.
