---
title: "Lab 2: Distances and Projections"
subtitle: 'ESS 523c: Environmental Data Science Applications'
author: 
  - name: Megan Hoover
  - email: megan.hoover21@gmail.com
format: html
---

### Lab Intro

This lab connects directly to recent policy discussions about border enforcement and the so-called “100-mile” border zone — for example, shifts in migration policy after the end of the pandemic-era Title 42 expulsions in 2023 have renewed attention on how proximity to the border affects enforcement, resource allocation, and civil liberties (see recent coverage: https://www.reuters.com/world/us/us-ceases-using-title-42-pandemic-order-control-migration-2023-05-11/).

More recent operational discussions involving ICE and border enforcement through 2025–2026 have emphasized how interior and border-zone enforcement strategies depend on distance-to-border metrics; readers can consult summaries on civil-liberties implications (ACLU) or policy analyses that discuss the “100-mile” enforcement footprint for background: https://www.aclu.org/know-your-rights/border-zone, https://www.americanimmigrationcouncil.org/blog/border-patrol-charlotte-atlanta-100-mile-zone/.

### Install Libraries

```{r}
pacman::p_load(
  tidyverse, sf, units, tigris, rnaturalearth, gghighlight, ggrepel, flextable
)

options(tigris_use_cache = TRUE)
```

### **Question 1: Build Your Spatial Dataset**

For this lab, we'll need three datasets:

1.  CONUS state boundaries (1.1)

2.  North American country boundaries — US, Mexico, Canada (1.2)

3.  All US cities (1.3)

#### **1.1 Define a Projection**

Distance calculations require a projected coordinate system that minimizes distance distortion at the scale of CONUS. We will use the **North America Equidistant Conic**:

```{r}
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```

This PROJ string defines:

| Parameter                | Value      | Meaning                       |
|--------------------------|------------|-------------------------------|
| `+proj=eqdc`             | —          | Equidistant Conic projection  |
| `+lat_0=40`              | 40°N       | Latitude of projection center |
| `+lon_0=-96`             | 96°W       | Central meridian              |
| `+lat_1=20`, `+lat_2=60` | 20°N, 60°N | Standard parallels            |
| `+datum=NAD83`           | —          | North American Datum 1983     |
| `+units=m`               | —          | Output in meters              |

Distortion is minimized between the two standard parallels — 20°N and 60°N encompasses all of CONUS, making this projection well-suited for national-scale distance analysis.

#### **1.2 CONUS State Boundaries**

Use `tigris::states()` to get US state boundaries, filter to CONUS, and transform to the equidistant projection.

**\
Tip**

`tigris` is the modern, CRAN-stable replacement for `USAboundaries`. It pulls boundaries directly from the US Census Bureau TIGER/Line files. Load the data with `states(cb = TRUE, progress_bar = FALSE)`. Filter the state codes (`STUSPS` column) to exclude Alaska, Hawaii, and US territories, then apply `st_transform(eqdc)` to reproject.

```{r}
?states
states<-states(cb = TRUE, progress_bar = FALSE) |>
  filter(!STUSPS %in% c("AK", "HI", "AS", "GU", "MP", "PR", "VI","DC")) |>
  st_transform(eqdc)
```

#### **1.3 North American Country Boundaries**

Use `rnaturalearth::ne_countries()` to get country boundaries for the United States, Canada, and Mexico.

**Tip**

`ne_countries()` returns an `sf` object directly when you specify `returnclass = "sf"`. Filter on the `admin` column to keep only the three countries you need, then transform to `eqdc`. See `?rnaturalearth::ne_countries` for available scales and parameters.

```{r}
?rnaturalearth::ne_countries
countries<- ne_countries(returnclass = "sf") |>
filter(admin %in% c("United States of America", "Canada","Mexico"))|>
 st_transform(eqdc) 
```

#### **1.4 US City Locations**

Go to [simplemaps.com/data/us-cities](https://simplemaps.com/data/us-cities) and download the free basic dataset into your `data/` directory.

Read it in with `readr::read_csv()` and explore it — understand what columns are available before proceeding.

The raw data is tabular — coordinates are stored as numeric columns, not geometry. Convert it to a spatial object using `st_as_sf()`, specifying which columns hold X and Y coordinates and what CRS they use. Raw lat/lon from a CSV is almost always WGS84 (EPSG:4326).

**Tip**

After creating the `sf` object, you’ll need to remove rows that fall outside CONUS. The `st_union(states)` geometry provides a boundary to filter against. See `?sf::st_filter` for how to select geometries that intersect or fall within a bounding polygon.

```{r}
?sf::st_filter
cities<- read_csv("data/uscities.csv") |>
  st_as_sf(coords = c("lng", "lat"), crs = 4326) |>
  st_transform(st_crs(states)) |> #Reprojects cities to same CRS as states object
   st_filter(st_union(states)) #Merges all state polygons into one big one & keeps only points intersecting boundary line.
  
```

**Verify all three datasets before proceeding**

Run these checks before moving to Q2 — a CRS mismatch here will silently corrupt every distance calculation downstream.

```{r}
# All three must return the same CRS
st_crs(states) == st_crs(countries)
st_crs(states) == st_crs(cities)


# Row counts should be reasonable
nrow(states)    # should be 48
nrow(cities)    # should be ~28,000

```

### **Question 2: Distance Calculations**

Here we calculate the distance from each US city to four boundaries:

1.  The US national border (coastline + land borders, resolved)

2.  The nearest state border (internal boundaries preserved)

3.  The Mexican border

4.  The Canadian border

In all cases, since we want distance to a **border** (a line), we must cast our polygon geometries to `MULTILINESTRING` before computing distances. The choice of `st_union` vs. `st_combine` determines whether interior boundaries are dissolved or preserved — a distinction that matters for questions 2.1 and 2.2.

**The geometry type determines the answer**

-   `POLYGON` / `MULTIPOLYGON` → distance to the interior is **0** (you’re inside the area)

-   `MULTILINESTRING` → distance to the **boundary line** — what we actually want

```         
# Wrong: distance to a polygon st_distance(city_point, state_polygon)   # returns 0 if city is inside  # Right: distance to the border st_distance(city_point, state_polygon |> st_cast("MULTILINESTRING"))
```

#### **2.1 Distance to US National Border (km)**

The national border is the **resolved** outer boundary of all states — internal state lines dissolved, only the coastline and land borders remain.

**Tip**

Use `st_union()` to merge all state polygons into a single boundary (dissolving internal borders), then cast to `MULTILINESTRING` so distances are measured to the **border line** rather than interior. Compute distance using `st_distance()` and convert the result to kilometers using `set_units()` from the `units` package — this ensures clean numeric output and correct unit handling downstream.

Store the result as a new column `dist_to_border` in your cities data frame. Produce a **flextable** of the 5 cities farthest from the national border — include city name, state, and distance.

```{r}
?st_distance
?set_units
us_border <- states |>
  st_union() |> # dissolve internal state boundaries
  st_cast("MULTILINESTRING")# convert polygon to boundary lines

cities <- cities |>
  mutate(dist_to_border = st_distance(geometry, us_border),
    dist_to_border = set_units(dist_to_border, "km"))


cities |>
  mutate(dist_to_border = drop_units(dist_to_border)) |> #removes the units, leaving a plain numeric column.
  arrange(desc(dist_to_border)) |>
  slice_head(n = 5) |>
  select(city, state_name, dist_to_border) |>
  flextable() |>
  set_header_labels(
    city = "City",
    state_name = "State",
    dist_to_border= "Distance to Border (km)") |>
  set_caption("Top 5 Cities Farthest From US Borders by")|>
  autofit()
```

#### **2.2 Distance to Nearest State Border (km)**

The nearest state border uses **preserved** internal boundaries — we want every state line, not just the outer edge. Use `st_combine()` which keeps all boundaries, then cast to `MULTILINESTRING`.

Store as `dist_to_state`. Produce a flextable of the 5 cities farthest from any state border.

```{r}
us_border <- states |>
  st_combine() |> # keep internal state boundaries
  st_cast("MULTILINESTRING")# convert polygon to boundary lines

cities <- cities |> #us_border is state border in this
  mutate(dist_to_state = st_distance(geometry, us_border),
    dist_to_state = set_units(dist_to_state, "km"))


cities |>
  mutate(dist_to_state = drop_units(dist_to_state)) |> #removes the units, leaving a plain numeric column.
  arrange(desc(dist_to_state)) |>
  slice_head(n = 5) |>
  select(city, state_name, dist_to_state) |>
  flextable() |>
  set_header_labels(
    city = "City",
    state_name = "State",
    dist_to_state= "Distance to State Border (km)") |>
  set_caption("Top 5 Cities Farthest From US State Borders")|>
  autofit()

```

**Tip**

Think carefully about why the answers for 2.1 and 2.2 differ. The city farthest from the national border might be close to a state line. The city farthest from any state line must be far from all internal AND external borders.

-   **2.1 (National border)**: distance only to the outer boundary (coastline + international borders).

    -    A city deep inside a state can be far from the national border but close to an internal state line.

<!-- -->

-    **2.2 (State borders)** → distance to any state line, including internal boundaries.

    -    The farthest city from all state lines must be roughly in the center of a large state.

#### **2.3 Distance to Mexico (km)**

Isolate Mexico from your countries object. Cast to `MULTILINESTRING`. Compute distance from each city to the Mexican border. Store as `dist_to_mexico`. Produce a flextable of the 5 cities farthest from Mexico.

```{r}
 notus<- countries |>
      filter(admin == "Mexico") |>
      st_cast("MULTILINESTRING") #convert polygon to boundary lines
  

cities <- cities |> #us_border is state border in this
  mutate(dist_to_mexico = st_distance(geometry, notus),
    dist_to_mexico = set_units(dist_to_mexico, "km"))


cities |>
  mutate(dist_to_mexico = drop_units(dist_to_mexico)) |> #removes the units, leaving a plain numeric column.
  arrange(desc(dist_to_mexico)) |>
  slice_head(n = 5) |>
  select(city, state_name, dist_to_mexico) |>
  flextable() |>
  set_header_labels(
    city = "City",
    state_name = "State",
    dist_to_mexico = "Distance to Mexico Border (km)") |>
  set_caption("Top 5 Cities Farthest From Mexico Border")|>
  autofit()

```

#### **2.4 Distance to Canada (km)**

Same approach as 2.3, for Canada. Store as `dist_to_canada`. Produce a flextable of the 5 cities farthest from Canada.

```{r}
notus<- countries |>
      filter(admin == "Canada") |>
      st_cast("MULTILINESTRING") #convert polygon to boundary lines
  

cities <- cities |> #us_border is state border in this
  mutate(dist_to_canada = st_distance(geometry, notus),
    dist_to_canada = set_units(dist_to_canada, "km"))


cities |>
  mutate(dist_to_canada = drop_units(dist_to_canada)) |> #removes the units, leaving a plain numeric column.
  arrange(desc(dist_to_canada)) |>
  slice_head(n = 5) |>
  select(city, state_name, dist_to_canada) |>
  flextable() |>
  set_header_labels(
    city = "City",
    state_name = "State",
    dist_to_canada = "Distance to Canadian Border (km)") |>
  set_caption("Top 5 Cities Farthest From Canadian Border")|>
  autofit()
```

### **Question 3: Mapping Distance Patterns**

Now visualize the distance data you computed. You will use `ggplot2` for all maps, `ggrepel::geom_label_repel()` for non-overlapping city labels, and `gghighlight` to emphasize subsets.

#### **3.1 Reference Map**

Create a base reference map showing:

-   North America (US, Canada, Mexico) as filled grey polygons

-   CONUS state boundaries as dashed lines

-   The 10 largest US cities by population as labeled points

**Tip**

Use `geom_sf()` for spatial layers. For labels that don’t overlap:

```         
ggrepel::geom_label_repel(   data = big_cities,   aes(geometry = geometry, label = city),   stat = "sf_coordinates",   size = 3 )
```

`stat = "sf_coordinates"` extracts point coordinates from the geometry for label placement — required when using `ggrepel` with `sf` objects.

```{r}
ggplot() +
  # North American countries
  geom_sf(data = countries, fill = "grey80", color = "black") +
   # CONUS state boundaries as dashed lines
  geom_sf(data = states, fill = NA, color = "black", linetype = "dashed") +
   # Plot and label the 10 largest cities inline
  geom_sf( data = cities %>% arrange(desc(population)) %>% slice_head(n = 10),
    color = "red", size = 2 ) +
  ggrepel::geom_label_repel(
    data = cities %>% arrange(desc(population)) %>% slice_head(n = 10),
    aes(geometry = geometry, label = city),
    stat = "sf_coordinates",
    size = 3,
    min.segment.length = 0 ) +
  # Theme and labels
  theme_minimal() +
  labs(
    title = "Reference Map: CONUS and Major US Cities",
    subtitle = "10 Largest US Cities by Population",
    caption = "Data: tigris, rnaturalearth, simplemaps")
```

#### **3.2 Distance to National Border**

Color all US cities by their distance to the national border using a continuous color scale. Highlight and label the 5 cities farthest from the border.

Your map should: - Use `scale_color_viridis_c()` for the distance gradient - Show the CONUS outline - Label the 5 farthest cities with `ggrepel`

```{r}
ggplot() +
  # CONUS outline
  geom_sf(data = states, fill = NA, color = "black") +
  # All US cities colored by distance to national border
  geom_sf(data = cities,
    aes(color = drop_units(dist_to_border)),
    size = 1) +
  scale_color_viridis_c(option = "plasma", name = "Distance to Border (km)") +
   # Label 5 farthest cities inline
  ggrepel::geom_label_repel(data = cities %>%
      mutate(dist_km = drop_units(dist_to_border)) %>%
      arrange(desc(dist_km)) %>%
      slice_head(n = 5),
    aes(geometry = geometry, label = city),
    stat = "sf_coordinates",
    size = 3,
    min.segment.length = 0) +
  # Theme and labels
  theme_minimal() +
  labs(
    title = "US Cities by Distance to National Border",
    subtitle = " Top 5 Farthest Cities Labeled",
    caption = "Data: tigris, rnaturalearth, simplemaps" )
```

In 2–3 sentences:

**(1) where are the cities farthest from the national border?** The cities farthest from the national border are located in Kansas.

**(2) Does the geographic pattern make sense given the shape of CONUS?** It makes sense because the projection is equidistant conic, which preserves distance.

**(3) Name one water resources application where distance to a national border would be a meaningful variable**. Distance of aquifers and watershed management to national borders matters in terms of how water is managed.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
